Extending Graphical Representations for Compact Closed Categories with Applications to Symbolic Quantum Computation

### 2008-05-04 papers/reasoning-graphically/paper.tex
final edits by me
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1481 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

Let $f: G \to H$ be a graph morphism and let $V' \subseteq V_G$. We say
that $f$ is \emph{strict for $V'$} if $\forall e \in E_H$, if $s_H(e)
\in f_V(V')$ or $t_H(e) \in f_V(V')$ then $\exists e' \in E_G$ such
that $f_E(e') = e$. Strictness ensures that there are no additional
edges connected to vertices in the image of $V'$.

@@

Given two graphs, $G$ and $H$, a pair of partial maps, $r_V: V_G
\rightharpoonup V_H$ and $r_E: E_G \rightharpoonup E_H$, is called a
\emph{partial graph morphism} if whenever $r_E(e)$ is defined then
$r_V(s_G(e))$ and $r_V(t_G(e))$ are defined, and the restriction of
$r_E$ to its preimage satisfies equations \eqref{eq:graph-hom1} and
\eqref{eq:graph-hom2} above.

@@

A substantial strand of work in quantum computation has involved the
development of high-level models of quantum processes based on compact
closed categories.  In these formalisms, initiated in
\cite{AbrCoe:CatSemQuant:2004}, quantum processes---such as quantum
logic gates, or the measurement of a qubit---correspond to arrows in a
category, while the different quantum data types, usually just arrays
of qubits, are the objects.  The graphical representation described in
the preceding section thus provides a very expressive notation quantum
processes.

@@

In terms of quantum processes, each edge in a graph represents a
qubit, although several edges may represent the same physical qubit at
different times.  An edge may even represent a ``virtual'' qubit which
stands for a correlation between different parts of the system.  The
maps $\delta_Z$ and $\epsilon_Z$ represent quantum operations which
respectively copy and delete the eigenstates of the Pauli $Z$
operator.\footnote{Uniform copying operations are forbidden by the
  no-cloning theorem \cite{no-cloning}, but such operations are
  possible if we demand only the eigenstates of some self-adjoint
  operator to be copied.  Other states will not not copied.  The same
  remarks hold true for erasing.}  Notice that $\delta_Z$ has one edge
in its domain representing the qubit to be copied, and two edges in
its codomain for the two copies it produces.  Similarly, $\epsilon_Z$
has one qubit as input and no outputs.  The adjoints $\delta^\dag_Z$
and $\epsilon^\dag_Z$ correspond to an operation known as
\emph{fusion}, and to the operation of preparing a fresh qubit in a
certain state.  The $\alpha_Z$ corresponds to phase shift of angle
$\alpha$ in the $Z$ direction.  The family of maps indexed by $X$ are
defined in exactly the same way, but relative to the Pauli $X$
operator rather than the $Z$.  All quantum operations may be defined
by combining these simple operations---which are essentially
classical---on two complementary observables.

\begin{remark}
  We emphasise that this is a mathematical notation is for
  representing quantum processes, unlike the so-called graph states
  used in measurement based quantum computation \cite{Briegel etc}.
  Their graphical notation is dual to ours, with the vertices
  representing qubits and the edges representing entanglement between
  the individual quibits.  Within the formalism of this paper, a state
  is simply a process with no inputs; that is, a concrete graph with
  empty domain.  An important application is to show that states or
  computations implemented differently are equivalent: this includes
  graph states, and measurement based computations more generally.
\end{remark}

@@

More formally, a !-box graph is a graph paired with a set !-boxes. Each
of these !-boxes is the set of nodes inside it and each !-box must be
disjoint from the others.\footnote{The disjointness condition avoids a
  slightly more complex, although also more expressive, semantics for
  overlapping node sets in the !-boxes. While such expressivity is
  interesting from a representational point of view, it is not needed
  to express the currently known rules for quantum computation.}  We
now define the concept of !-box matching which we then use to provide
a formal semantics for the !-boxed graphs.

%\textbf{Matching.} 
To formalise the intuitive notion that a !-box represents arbitrary
number of copies of the subgraph made from its nodes, we introduce
\emph{!-box matching}. This is a binary relation on graphs, written
infix as $G \leq_! H$ when $G$ matches $H$. The set of !-box graphs
matching $G$ is defined as the set of graphs closed by the following
operations on $G$:

\begin{description}
\item[copy]: copies a !-box, $b$ in a graph to produce a new graph
  with two copies of the !-boxed subgraph, $b$ is the old one and $b'$
  is the new one. The set of !-boxes in the copied graph now also
  contains the new !-box $b'$. Any edges between a node, $n$, inside
  the !-box $b$, and a node, $m$, outside it, get copied so that there
  is a new edge from $m$ to the new copy of $n$ in $b'$.

@@

We give a formal semantics to !-box graphs by defining them in terms
of a set of graphs in the underlying representation. We denote the
interpretation of a !-box graph $G$ by $\binterp{G}$ and say that
members of $\binterp{G}$ are instances of $G$.

\begin{definition}
  \emph{Interpretation of !-box Graphs}: $\binterp{G}$ is the subset
  of graphs matched by the !-box graph that have no !-boxes.
\end{definition}

@@

\begin{theorem}
\label{thm:bang-box-respect}
\emph{!-Matching respects !-box semantics}: $G \leq_! H
\Leftrightarrow \binterp{G} \supseteq \binterp{H}$. The proof is a
simple consequence from the definition of $\binterp{G}$ being a subset
of the graphs that match $G$.
\end{theorem}

@@

For the purposes of this paper, the underlying graph representation of
!-boxed graphs is variable-node graphs. This gives rise to graph
patterns which we now discuss in more detail. 
% The set of variable-node
% graphs for a graph $G$ with !-boxes, is denoted by $\binterp{G}$.

@@

Our model of quantum computation requires no data for the edges. The
nodes on the other hand are either $H$ (a Hadamard gate), with no
additional data, or have an \emph{angle} and are either the $X$ or $Z$
transforms. For their part, angles are expressed as rational numbers
which correspond to the coefficient of $\pi$ in the underlying matrix.

@@

>>>>

%%% TODO removed this because we don't need partial maps
%%% unless we are talking about plugging.
%%%
% Given two graphs, $G$ and $H$, a pair of partial maps, $r_V: V_G
% \rightharpoonup V_H$ and $r_E: E_G \rightharpoonup E_H$, is called a
% \emph{partial graph morphism} if whenever $r_E(e)$ is defined then
% $r_V(s_G(e))$ and $r_V(t_G(e))$ are defined, and the restriction of
% $r_E$ to its preimage satisfies equations \eqref{eq:graph-hom1} and
% \eqref{eq:graph-hom2} above.

\begin{definition}
  Let $f: G \to H$ be a graph morphism and let $V' \subseteq V_G$. We
  say that $f$ is \emph{strict for $V'$} if $\forall e \in E_H$, if
  $s_H(e) \in f_V(V')$ or $t_H(e) \in f_V(V')$ then $\exists e' \in
  E_G$ such that $f_E(e') = e$.
\end{definition}
\noindent Strictness ensures that there are no additional
edges connected to vertices in the image of $V'$.

@@

@@

A substantial strand of work in quantum computation has involved the
development of high-level models of quantum processes based on compact
closed categories.  In these formalisms, initiated in
\cite{AbrCoe:CatSemQuant:2004}, quantum processes---such as quantum
logic gates, or the measurement of a qubit---correspond to arrows in a
category, while the different quantum data types, usually just arrays
of qubits, are the objects.  The graphical representation described in
the preceding section thus provides a very expressive notation for quantum
processes.

@@

In terms of quantum processes, each edge in a graph represents a
qubit, although several edges may represent the same physical qubit at
different times.  An edge may even represent a ``virtual'' qubit which
stands for a correlation between different parts of the system.  The
maps $\delta_Z$ and $\epsilon_Z$ represent quantum operations which
respectively copy and delete the eigenstates of the Pauli $Z$
operator.\footnote{Uniform copying operations are forbidden by the
  no-cloning theorem \cite{Wootters1982A-single-quantu}, but such operations are
  possible if we demand only the eigenstates of some self-adjoint
  operator to be copied.  Other states will not not copied.  The same
  remarks hold true for erasing \cite{Pati2000Impossibility-o}.}  Notice that $\delta_Z$ has one edge
in its domain representing the qubit to be copied, and two edges in
its codomain for the two copies it produces.  Similarly, $\epsilon_Z$
has one qubit as input and no outputs.  The adjoints $\delta^\dag_Z$
and $\epsilon^\dag_Z$ correspond to an operation known as
\emph{fusion}, and to the operation of preparing a fresh qubit in a
certain state.  The $\alpha_Z$ corresponds to phase shift of angle
$\alpha$ in the $Z$ direction.  The family of maps indexed by $X$ are
defined in exactly the same way, but relative to the Pauli $X$
operator rather than the $Z$.  All quantum operations may be defined
by combining these simple operations---which are essentially
classical---on two complementary observables.

We emphasise that this is a notation is for representing quantum
processes, rather than quantum states.  In this setting a state is
simply a process with no inputs; that is, a concrete graph with
empty domain.
  Since our formalism is based on the
underlying mathematical structure rather than any particular model of
quantum computation, it is capable of representing  quantum circuits,
and measurement-based quantum computations, among other models.
Indeed, an important application is to show that states or 
computations implemented differently are equivalent.

@@

More formally, a \emph{!-box graph} is a pair $(G, {\cal B})$ where $G$ is a
graph and $\cal B$ is a set of subsets of $V$ such that  whenever
$b_1, b_2 \in {\cal B}$ then $b_1 \cap b_2 =\emptyset$,\footnote{One
  could consider more expressive notions of nested, or overlapping,
  node sets in the !-boxes. While such expressivity is 
  interesting,  it is not required for the system we formalise here.}  and
subject to the further restriction that no edge may connect vertices
in distinct !-boxes, i.e. if $v_1 \in b_1$, $v_2 \in b_2$ and $s(e) =
t(e)$ then $b_1 = b_2$.  
To formalise the intuitive notion that a !-box represents arbitrary
number of copies of the subgraph made from its nodes, we introduce
\emph{!-box matching}.   This binary relation, written infix as
$\leq_!$, is defined such that 
$(G,{\cal B}) \leq_! (H, {\cal C})$ whenever $(H, {\cal C})$ can be
obtained from $(G,{\cal B})$ be a sequence the following operations:

%
% TODO there ought to be formal definitions of these but not enough
% space in this version
\begin{description}
\item[copy]: copies a !-box, $b$ in a graph to produce a new graph
  with two copies of the !-boxed subgraph, $b$ is the old one and $b'$
  is the new one. The set of !-boxes in the copied graph now also
  contains the new !-box $b'$. Any edges between a node, $n$, inside
  the !-box $b$, and a node, $m$, outside it, get copied so that there
  is a new edge from $m$ to the new copy of $n$ in $b'$.

@@

We give a formal semantics to !-box graphs by defining them in terms
of a set of graphs in the underlying representation. We denote the
interpretation of a !-box graph $(G,{\cal B}$ by $\binterp{(G,{\cal
    B})}$ and say that members of $\binterp{G}$ are instances of $G$.

\begin{definition}
  The \emph{interpretation} $\binterp{(G,{\cal B})}$ is a
  set of concrete graphs defined 
  \[
  \binterp{(G,{\cal B})} = \{ H \;\;|\;\; (G,{\cal B}) \leq_! (H,\emptyset) \}
  \]
  i.e. the those graphs matched by the !-box graph that have no
  !-boxes.
\end{definition}

@@

\begin{proposition}
\label{thm:bang-box-respect}
\emph{!-Matching respects !-box semantics}: $G \leq_! H
\Leftrightarrow \binterp{G} \supseteq \binterp{H}$. The proof is a
simple consequence from the definition of $\binterp{G}$ being a subset
of the graphs that match $G$.
\end{proposition}

@@

For the purposes of this paper, the underlying graph representation of
!-boxed graphs is variable-node graphs. This gives rise to \emph{graph
patterns} which we now discuss in more detail. 
% The set of variable-node
% graphs for a graph $G$ with !-boxes, is denoted by $\binterp{G}$.

@@

Our model of quantum computation requires no data for the edges. The
nodes on the other hand are either $H$ (a Hadamard gate), with no
additional data, or have an \emph{angle} and a \emph{colour}, expressing that
they are defined in the $X$ or $Z$ basis. For their part, angles are
expressed as rational numbers 
which correspond to the coefficient of $\pi$ in the underlying matrix.

@@

We note that our graphical notation has little connection
to \emph{graph states} as used in various approaches to
measurement-based quantum computation \cite{Raussendorf-2001}.  In
that approach the graph structure is used to provide a description of
the entanglement in a state:  it does not provide a complete
description of a computation.
